---
title: "ToothGrowth Project"
output: pdf_document
---

We analyze the ToothGrowth data in the R datasets package.

### Load the ToothGrowth data and perform some basic exploratory data analyses.



*From the R datasets package ...*

"The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC))."


We load the libraries that we'll use and the ToothGrowth data set.

```{r, warning=FALSE}
library(datasets)
library(ggplot2)

data(ToothGrowth)
```
Let's look at what's inside.

```{r}
str(ToothGrowth)
```

So we have 60 observations (i.e. rows) of 3 variables (i.e. columns).

### Provide a basic summary of the data.

Let's take a quick summary of the data.

```{r, warning=FALSE}
summary(ToothGrowth)
```




Build a contingency table of the counts at each combination of factor levels.
```{r, warning=FALSE}
table(ToothGrowth$dose, ToothGrowth$supp)
```

Let's look at box plots for tooth growth as a function of dose and as a function of delivery method.

```{r, warning=FALSE}
dose_box <- ggplot(data=ToothGrowth, aes(x=as.factor(dose), y=len)) + 
            xlab("Dose") + ylab("Length") + theme(legend.title=element_blank()) +
            ggtitle("Box Plot for Tooth Length as a Function of Dose") +
            geom_boxplot(aes(fill=factor(dose))) + theme_bw()


dose_box
```

<br/> From the box plot above, it appears that a greater dose a greater effect on tooth length.
<br/>
<br/>



```{r, warning=FALSE}
supp_box <- ggplot(data=ToothGrowth, aes(x=as.factor(supp), y=len)) + 
            xlab("Delivery Method") + ylab("Length") + theme(legend.title=element_blank()) +
            ggtitle("Box Plot for Tooth Length as a Function of Supplement Type") +
            geom_boxplot(aes(fill=factor(supp))) + theme_bw()


supp_box
```
<br/> From the box plot above, it appears that orange juice *might* have a slightly greater effect on tooth length, than vitamin C.
<br/>
<br/>

### Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. 
(Only use the techniques from class, even if there's other approaches worth considering)

We use t-tests for doses and for supplements.

First, we look at doses.


```{r}
# t-test for different dosages
dose_0.5 <- subset(ToothGrowth, dose == 0.5)
dose_1 <- subset(ToothGrowth, dose == 1)
dose_2 <- subset(ToothGrowth, dose == 2)

```
We consider three cases:

```{r}
t.test(dose_0.5$len,dose_2$len)
```

```{r}
t.test(dose_1$len,dose_2$len)
```

```{r}
t.test(dose_0.5$len,dose_1$len)
```

Looking at the t-test results for the differing doses, we see small p-values. For the difference between a dosage of 0.5 and 2, we have a p-value < 0.00000000001. The ecidence is strong, that increased doses results in increased tooth length.


<br/>
Next, we look at supplement type.

```{r}
# t-test for supplement types
supp_OJ <- subset(ToothGrowth, supp == "OJ")
supp_VC <- subset(ToothGrowth, supp == "VC")
```
There is only one case to consider:

```{r}
t.test(supp_OJ$len,supp_VC$len)
```

We have a p-value equal to 0.06063 which is greater than the 0.05 significance level.
This leads us to conclude that the effect of the supplements is similar.
<br/>

### State your conclusions and the assumptions needed for your conclusions. 

We conclude that supplement types have similar effects on tooth growth.
On the other hand, increasing doses leads to a significant increase in tooth length.

The assumptions underlying this analysis are that the guinea pigs were sampled from a normally distributed population. The samples were drawn randomly.
